---
title: "review"
output:
  rmarkdown::md_document:
    pandoc_args: [
      "--wrap=none"
    ]
---

## Package Review

*Please check off boxes as applicable, and elaborate in comments below.  Your review is not limited to these topics, as described in the reviewer guide*

- **Briefly describe any working relationship you have (had) with the package authors.**
- [X] As the reviewer I confirm that there are no [conflicts of interest](https://devguide.ropensci.org/policies.html#coi) for me to review this work (if you are unsure whether you are in conflict, please speak to your editor _before_ starting your review).

#### Documentation

The package includes all the following forms of documentation:

- [ ] **A statement of need:** clearly stating problems the software is designed to solve and its target audience in README

The opening paragraphs of the README are good, and I think that this R package solves a challenging problem, so firstly, well done! I think could be made a little bit clearer in terms of the problem it solves, and the input it takes. While I find the photos useful, it initially made me think that this software takes images as input. I would suggest something more like what is in the vignette to start:

> `fellingdateR` offers a set of functions that assist in inferring felling date estimates from dated tree-ring series.

Then, describe the problem you want to solve, which I think is estimating _when_ the timber was cut down. Then show the data, explain what the columns mean, and how this might be a typical example of dated tree-ring series data.

Then show a short example of the output, clearly demonstrating the problem the package solves.

The rest of the first paragraph:

> The presence of (partially) preserved sapwood or waney edge allows to estimate a range for the actual felling date, for individual series as well as for a group of timbers. Furthermore, an additional function provides a tool to sum sapwood probability distributions, comparable to ‘summed probability densities’ commonly applied to sets of radiocarbon (14C) dates.

Is important, but I think could go into more of a methods/general introduction part of the README, perhaps further down.

I'm not sure what the images show me, and so to communicate this effectively I think they should contain a caption.

I think the target audience could be more clearly stated in the README. Perhaps at the end of the first paragraph.

- [X] **Installation instructions:** for the development version of package and any non-standard dependencies in README

All installed well for me!

- [X] **Vignette(s):** demonstrating major functionality that runs successfully locally

It did run successfully locally! `T` and `F` should be specified as `TRUE` and `FALSE`.

- [X] **Function Documentation:** for all exported functions
- [X] **Examples:** (that run successfully locally) for all exported functions

The examples ran without error, using:

```r
devtools::run_examples()
```

- [ ] **Community guidelines:** including contribution guidelines in the README or CONTRIBUTING, and DESCRIPTION with `URL`, `BugReports` and `Maintainer` (which may be autogenerated via `Authors@R`).

There are no community guidelines in the README, I see them in the file: `.github/CONTRIBUTING.md`, but these are not linked to in the README. Once these are linked, e.g., by writing something like:

```

## Code of Conduct

Please note that the visdat project is released with a [Contributor Code of Conduct](https://github.com/hanecakr/fellingdateR/blob/main/.github/CONTRIBUTING.md). By contributing to this project, you agree to abide by its terms.

```

#### Functionality

- [X] **Installation:** Installation succeeds as documented.
- [X] **Functionality:** Any functional claims of the software been confirmed.
- [X] **Performance:** Any performance claims of the software been confirmed.
- [X] **Automated tests:** Unit tests cover essential functions of the package and a reasonable range of inputs and conditions. All tests pass on the local machine.

All tests pass - unit tests seem quite good coverage, evaluated using `devtools::test_coverage()`.

- [ ] **Packaging guidelines**: The package conforms to the rOpenSci packaging guidelines.

- package name passes checks on `available::available("fellingdateR")`
- I think if possible the author should consider renaming the package to all lowercase, `fellingdater` or `fellingdatr`.
- There are other considerations that I think mean it does not currently conform to the rOpenSci packaging guidelines. Rather than discuss them in too much depth here, I will put them in the review section below.

Estimated hours spent reviewing: 5

- [X] Should the author(s) deem it appropriate, I agree to be acknowledged as a package reviewer ("rev" role) in the package DESCRIPTION file.

---

### Review Comments

I wanted to open by saying that while I have a lot of feedback, I think that this is a great piece of software that helps solve a tough problem, so well done on the author for writing this! I hope that the feedback is useful 😄 . Please let me know if something is not clear or if you need help implementing these, or further information. Thank you for submitting this software, I enjoyed reviewing it.

#### General comments

There are a fair few examples from the [rOpenSci packaging guide](https://devguide.ropensci.org/building.html), which I don't think are followed, I have gone through the guide and written some examples here. After the author makes these changes, I would recommend they double check the guide.

- Recommend making sure all functions and objects use snake_case. 
- argument name uses `x` for most data frame inputs. I would recommend considering naming data things `data` or `.data` or similar to help distinguish them from a vector, `x`. Not required but worth considering, I think.
- There is some use of `cat` in the package, recommend using `cli` as described in the [tidyverse style guide on writing error messages](https://style.tidyverse.org/error-messages.html). I expand on this below.
- Code style is not consistent, there is mixed use of the number of indentations : between 0 and 8 spaces. I would recommend applying the tidyverse style guide to the package with `styler::style_pkg()`
- Indenting code is important but this 8 space indentation is not consistent with other indentation used in your package, and when reading the code gives the impression that the code is happening inside some/several if/else/for control statements. I would recommend applying a style guide such as the tidyverse style guide or similar, to the code, so that indentation is consistent.
- Due to indentation, a lot of lines of code go over 80 characters. I think it is worth the time to re-indent, or rewrite some code by using explaining variables, so the code doesn't go over 80 characters
- `=` is sometimes used over `<-` - I recommend using `<-` consistently.
- There is no top level documentation for `?fellingdateR` - this could be achieved using `usethis::use_package_doc()`.
- package should use a website. See the [ropensci guide on building a website](https://devguide.ropensci.org/building.html#website)
- internal functions, like `d.dens` and `d.count` should have a `#' @noRd` tag to mark is as an internal function
- examples in code should use all argument parameters
- Recommend the author reads through the [CRAN gotchas](https://devguide.ropensci.org/building.html#crangotchas)
- What does the `sw` stand for in things like `sw_combine` and co?
- Some of the documentation uses reversed backticks, which I haven't seen before, e.g.: `´n_sapwood´ and ´count´`

- There are still a few lines of code that don't pass the `goodpractice::gp()` checks. In particular, I think these comments are important:

```
  ✖ write short and simple
    functions. These functions
    have high cyclomatic
    complexity (>50): read_fh
    (150). You can make them
    easier to reason about by
    encapsulating distinct steps
    of your function into
    subfunctions.
  ✖ use '<-' for
    assignment instead of '='.
    '<-' is the standard, and R
    users and developers are used
    it and it is easier to read
    your code for them if you use
    '<-'.
  ✖ avoid long code lines,
    it is bad for readability.
    Also, many people prefer
    editor windows that are about
    80 characters wide. Try make
    your lines shorter than 80
    characters
  ✖ avoid sapply(), it is
    not type safe. It might return
    a vector, or a list, depending
    on the input data. Consider
    using vapply() instead.
  ✖ avoid 1:length(...),
    1:nrow(...), 1:ncol(...),
    1:NROW(...) and 1:NCOL(...)
    expressions. They are error
    prone and result 1:0 if the
    expression on the right hand
    side is zero. Use seq_len() or
    seq_along() instead.
  ✖ avoid 'T' and 'F', as
    they are just variables which
    are set to the logicals 'TRUE'
    and 'FALSE' by default, but
    are not reserved words and
    hence can be overwritten by
    the user.  Hence, one should
    always use 'TRUE' and 'FALSE'
    for the logicals.
```

- You interchange between using `=` and `<-` in your code. I would recommend using `<-` only. See for example in `cor_table.R`:

```r
if (is.null(y)) {
     y = x
     noRef = TRUE
}
else {
     noRef = FALSE
     y_ori <- y
}
```


- Error messages. I would recommend building input checking functions to assist in how your write up error messages. There are a few key benefits to this:

  - The input checking function then does not get in the way of understanding the intent of your function code. 
  - You can reuse the input checking functions, so you don't need to write them again.
  - using `cli` to build the error messages allows you to use `glue` strings, so you don't have to try and quote or inject other information into the message string, it should be easier to add details you care about.
  
- Error messages are hard to write well, and it's great that you've included some good input checking! I think you could make the error functions a bit better for the user by following the [tidyverse style guide on error messages](https://style.tidyverse.org/error-messages.html).

- explaining variables. I've mention this a few times in the other functions in your package, I think it would be worthwhile searching through your cases of using `if` and if there is a long conditional in there, e.g., 

`any(pdf_matrix[, 2:length(keycodes) + 1] == 1, na.rm = TRUE))`

- Then I think it would be worthwhile either writing a small wrapper function to identify this, or wrap that up in an explaining variable.

- `plot = TRUE` as a function option.

  - I believe plotting functions should be separate to statistical transformation/operations. You have written these in `ggplot`, and so you can specify an `autoplot` method or a separate `plot_<function>` command.
  - The user should be able to reconstruct the plot from the data that you give them in these functions. E.g., they should be able to get the key information out, such as for `sw_interval`, the following information should be given from the function: `n`, `hdi`, and the number of sapwood rings.
  - It is not clear to me how to get this information, and I think that this is really important that the user doesn't end up locked into a plot to get their vital statistics. If they want to be able to generate tables or other statistics, then they cannot do this programmatically, they would have to physically eyeball a plot and record down the numbers, like `hdi (95.4%) = between 8 and 26 sapwood rings". Which is prone to errors.

- consistent file names. Some of the files have camelCase names (`movAv.R`), others are snake_case. I would recommend sticking to a consistent naming scheme, snake_case.
- I would try and avoid having `else` statements contain errors/stops/warnings/messages. This is because in order to understand the message at the end, you need to then walk back up through the condition of logic beforehand. The way to avoid this is to clearly state the error condition at the top.


##### Input checking

I would recommend writing small helpers for input checking, and considering using `cli` to help write error messages, as it means you could transform this:

```r
if (!inherits(x, "rwl"))  {
     warning("'x' is not class rwl")
}
if (!inherits(y, "rwl"))  {
     warning("'y' is not class rwl")
}
```

Into:

```r
warn_if_not_rwl(x)
warn_if_not_rwl(y)
```

And that code could look like this:

```{r}
warn_if_not_rwl <- function(x,
                            arg = rlang::caller_arg(x),
                            call = rlang::caller_env()){
     cli::cli_warn(
          c("{arg}' is not of class {.cls rwl}")
     )
}
```

Similarly, 

```r
increasing_consecutive_years <- all(diff(as.numeric(row.names(x))) == 1)
if (!increasing_consecutive_years) {
     stop(
          "The tree-ring series 'x' have/has no consecutive years in increasing order as rownames."
     )
}
```

Could be written as a function:

```r
check_if_increasing_consecutive_years(x)
check_if_increasing_consecutive_years(y)
```

Admittedly, I do have a strong preference for writing these types of functions,
[having written about it recently](https://www.njtierney.com/post/2023/12/06/long-errors-smell/), but I do think that at least using explaining variables, which you've already done in places like:

```r
increasing_consecutive_years <- all(diff(as.numeric(row.names(x))) == 1)
```

Are a great idea, and there are a few notable places where that would help 
make the code a bit easier to read, e.g., 

```r
any(
length(min_overlap) != 1 |
!is.numeric(min_overlap) |
min_overlap %% 1 != 0 |
min_overlap < 3
)
```


#### cor_table.R

Refactoring `values` argument of `cor_table`. There is a lot of input checking for the `values` argument. I think that things such as :

```r
if ("glk" %in%  values) {
```

And so on indicate to me that these could be written up as separate functions, which could return a list of their inputs, perhaps. These could then be delivered using `switch`, which I often forget how to use, but it would be something like:

```r
values_output <- switch(values,
       "glk" = values_glk(inputs),
       "pearson" = values_pearson(inputs))
```

Examples should demonstrate all types of the inputs for the function arguments.

#### data.R

I would recommend standardising the dataset names to be all lowercase, so that they are easier to remember. E.g., `Sohar_2012_FWE_c` becomes: `sohar_2012_fwe_c`

#### fd_report.R

I think that `fd_report` could be renamed `felling_report` or `felling_date_report` or similar. While `fd` is concise, I think it doesn't help facilitate discoverability of the functions. 

Similar to `cor_table.R`, I think that:

```r
if (!series %in% names(df)) {
      stop("--> 'series' does not exist")
}
if (!last %in% names(df)) {
      stop("--> 'last' does not exist")
}
if (!n_sapwood %in% names(df)) {
      stop("--> 'n_sapwood' does not exist")
}
if (!waneyedge %in% names(df)) {
      stop("--> 'waneyedge' does not exist")
}
```

Could be rewritten as `check_if_variable_exists()`. Something like:

```{r}
#| error: true
check_if_variable_exists <- function(x,
                                     df,
                                     arg = rlang::caller_arg(x),
                                     call = rlang::caller_env()){
     arg_in_data <- x %in% names(df)
     if (!arg_in_data) {
          cli::cli_abort(
               c("{.arg {arg}} does not exist")
          )
     }
}

example_checker <- function(x, 
                            series = "series", 
                            last = "last"){
     check_if_variable_exists(series, x)
     check_if_variable_exists(last, x)
}

example_checker(mtcars, 
                series = "wrong")

```

#### get_header.R

This function should move the `cat` message up the top - and should not use `cat`, instead using one of the `cli` functions, like `cli_abort`.

I think you could use `structure` instead of setting attributes to NULL:

```r
attr(rwl, "row.names") <- NULL
attr(rwl, "po") <- NULL
attr(rwl, "class") <- NULL
attr(rwl, "names") <- NULL

## becomes

rwl <- structure(
          rwl,
          row.names = NULL,
          po = NULL,
          class = NULL,
          names = NULL
     )
```

Although I think that they are functionally the same, so feel free to ignore!

#### hdi

This function uses `=` and `<-` - suggest sticking to just `<-`

#### movAv

I think this starting chunk would be clearer if only `if` and not `else` is used.

The stop error can move to the top of this, so we clearly capture if `align` is not "center" or "right" or "left". This makes it easier to understand the conditions of error.

```r
if (align == "center") {
     before <- floor((w - 1) / 2)
     after  <- ceiling((w - 1) / 2)
} else if (align == "right") {
     before <- w - 1
     after  <- 0
} else if (align == "left") {
     before <- 0
     after  <- w - 1
} else {
     stop("'align' should be 'center', 'left' or 'right'")
}
```

I suggest using another explaining variable inside `mean`:

```r
mean(x[max(0, (i - before)):(i + after)], na.rm = TRUE)

## to something like:

earliest_to_latest <- x[max(0, (i - before)):(i + after)]
mean(earliest_to_latest, na.rm = TRUE)

## or given that this is repeated later
## potentially write this up as a function for reuse?
mean_earliest_latest(x, i, before, after)
```

As that `mean` statement is a bit involved to unfurl.

Similarly, the pattern, `if (edges == "fill") {` and `} else if (edges == "nofill") {` should be bundled up into a function and applied with `switch`

#### read_fh.R

- Nice work in the attribution to the other previous work this extends. It looks like this is borrowed from `dplR` directly, and as such there are small style changes. I think it is worthwhile updating the code style to fit within your package.
- Be consistent with naming variables, `header.taken` should be `header_taken` etc.
- There are a few random comments that I'm not sure need to be there:

```
        # NEW: verbose = TRUE, header = FALSE
        inp <- readLines(fname, ok = TRUE, warn = FALSE)
        # NEW: removes empty lines in .fh file
        inp <- inp[nchar(inp) != 0]
        ## Get start and end positions of headers and data blocks
        header.begin <- grep("^HEADER:$", inp)
        # NEW: Quadro => chrono
        # NEW: Double => half chrono
lengths <- numeric(n) # commit Ronald Visser
```

I have found that moving comments either into documentation or into issues to help track them is helpful, but I appreciate that sometimes it is best to leave them in the code, but just something that might be worth thinking about :)

Tidying up the error messages in this function would make some of these nested if/else clauses easier to understand.

This is a pretty massive function, a bit over 1200 lines of code. I would recommend breaking down the steps inside this into smaller functions, as this will make the code easier to reason with and maintain in the future.

#### sw_combin_plot.R

This is the first time I've seen `############` comment blocks - I'm all for stylistic choices but I am not sure this is needed, especially if this isn't used in other functions.

I've not seen this pattern to avoid R CMD Check notes before

```{r}
   # to avoid notes in CMD check
   year <-
      p <-
      lower <-
      upper <- COMB <- last <- n_sapwood <- A_i <- agreement <- NULL
```

My tactic has always been to have a separate definition of these, as answered by [Carson Sievert on the posit community paage](https://community.rstudio.com/t/how-to-solve-no-visible-binding-for-global-variable-note/28887). I don't think there's anything inherently wrong with that, but I could imagine that in some cases this could accidentally erase inputs. Something to be aware of, perhaps?

I am all for using the new base R pipe `|>` - however you need to update your Depends in your DESCRIPTION like so in order to use it, since it only came out in R 4.1.0:

```
Depends: 
    R (>= 4.1.0)
```

This comment should probably live in a github issue or just be removed:

```r
      # NEXT LINE TRIGGERS WARNING
      # Warning message:
      # Using one column matrices in `filter()` was deprecated in dplyr 1.1.0.
      # ℹ Please use one dimensional logical vectors instead.
      # ℹ The deprecated feature was likely used in the fellingdateR package.
      # Please report the issue to the authors.
      # { if (nrow(summary |> dplyr::filter(agreement == "poor")) != 0)
      # replaced by:
```


#### sw_combine.R

This error should check each of the conditions separately - either it has missing values, or it is not numeric.

```r
if (any(is.na(endDate)) | !is.numeric(endDate)) {
     stop(
          "--> Please check the column with 'end dates'.
Some values are possibly missing or the values are not numeric"
     )
}
```

#### sw_data_info.R

I think these error messages would benefit from using `cli`, as discussed above.

#### sw_data_overview.R

This is a nice function to include to facilitate data discovery

#### sw_interval_plot.R

This code 

```r
if (all(
     !(attributes(x)$names) %in% c(
          "year",
          "n_sapwood",
          "p")
))
     stop("Input differs from output sw_interval()")
```

Could be rewritten as an error function or the condition in `if` could be expressed as a function.

#### sw_interval.R

In the final line of documentation for this function there is a hanging sentence:

```r
#' @return Depends on the value of `hdi`.
#'
#'  * If `hdi = TRUE`, a `numeric vector` reporting the upper and lower limit
#'   of the hdi (attributes provide more detail on `credMass` and the applied
#'   sapwood model (`sw_data`)).
#'  * If `hdi = FALSE`, a `matrix` with scaled p values for each number of
#'   observed sapwood rings. This matrix
```

#### sw_model.R

Great to see input checking at the top of the function - I do think these should be rewritten as check input functions.

Helper function `d.count` I think should be put into a separate R file called `utils.R` or `helpers.R`

`d.count` should use `switch` pattern and pass functions rather than using `if` controls.

`d.count` should be `d_count`

#### sw_sum_plot.R

indentation in this code is not consistent - recommend applying a style guide.

Examples should show different variations possible for function arguments. E.g., `bar_col`, `spline_col`, `dot_col`, and `dot_size` should all be specified in the examples so the user can see what the input should/could be.

#### sw_sum.R

See note above on including plots.

#### tests

- Do not need to namespace testthat calls, e.g., remove `testthat::`
- consider using snapshot testing, to capture exact values and shape of data that should be stable - rather than always testing for data shape and type columns, this should be able to capture those outputs
- consider snapshot error testing to capture exact error messages
- consider using `vdiffr` for testing ggplot plots. See [visdat](https://github.com/ropensci/visdat/blob/master/tests/testthat/test-vis-dat.R) for examples
